 

### 2021-08-09 vertical-thermal-review.Rmd
added citations for modelling

**The findings of this review reinforce the notion that representing vertical structuring is essential to capturing forest dynamics under global change.** Improved representation of vertical variation in forest canopy architecture and ecosystem function is critical for representing thermal sensitivity and has repeatedly been identified as an important step toward reduced uncertainty and more accurate characterization of biologically mediated feedbacks (Moorcroft, Hurtt, & Pacala, 2001; Banerjee & Linn 2018; Bonan et al. 2021). In addition, this review finds that, although larger trees face disproportionate stress during drought, understory trees have fewer cooling mechanisms and may also be stifled under hot, humid conditions. **A key question is whether existing models adequately represent the processes that underpin understory tree responses to thermal stress as well as large tree responses.** Most models have been developed to capture dynamics in the canopy, or overstory, given the disproportionate role of these large trees in ecosystem dynamics. Less attention has focused on developing and validating understory tree dynamics and responses to perturbations in models. This is likely due in large part to the fact that observations are only beginning to shed light on the vulnerability of understory trees to changes in temperature and other abiotic conditions [e.g., @rollinsonClimateSensitivityUnderstory2020].

Though an improvement over big-leaf models, DGVMs that separate the canopy into only two layers (e.g., sunlit and shaded portions) may not be able to capture important within-canopy variation in terms of leaf dynamics (e.g., seasonal shifts in vertical leaf area distributions, Smith et al. 2019) and functions (e.g., thermal responses, as we present in this paper). At the very least, multi-layered ecosystem models will likely be necessary for accurately predicting future forest function (e.g., de Pury & Farquhar, 1997; Bonan et al 2021). In addition, capturing the vertical gradients in NPP, GPP, and respiration, and subsequently the net ecosystem effects, requires improved characterization of the functional response of leaf-level processes to vertically varying abiotic conditions, and the role of traits and abiotic conditions in mediating responses to thermal sensitivity.

@@

Detailed structural information from lidar data has been used to measure tree height, DBH, and crown dimensions [@fisherECOSTRESSNASANext2020], evaluate spatial variation in vertical leaf area density profiles [@dettoSpatialVariabilityTropical2015], and resolve complex seasonal and diurnal variation in shortwave radiation forcing in ecosystems [@musselmanEstimationSolarDirect2013]. In addition, shading by the overstory can be better represented to predict understory temperatures by modeling the time-varying interaction of the forest canopy with solar radiation, as has been done in other LIDAR-based analyses (e.g., Davis et al. 2019). Satellite and airborne thermal infrared remote sensing analyses are now being conducted at regional and continental scales, although very few applications exist at finer-scales necessary for understanding of vertical variation in plant canopy temperatures (Johnston et al. in review). (Pau et al. 2018) used data from a tower-based FLIR camera in combination with eddy-covariance data and found that tropical forest canopy temperatures were more strongly associated with GPP than air temperatures or VPD. In a savanna system in the western U.S., (Johnston et al. in review) found significant vertical variation in plant temperature, with warmer temperatures in the understory compared to the top of the blue oak canopies. This finding was the opposite of what was expected and occurred in association with very high grass temperatures. The growing availability of these data makes this an opportune time to link fine-scale and landscape-scale measurements with the type of in situ measurements reviewed above to further explore ecosystem-scale patterns in vertical temperature gradients seasonally and across biomes.

>>>>

**The findings of this review reinforce the notion that representing vertical structuring is essential to capturing forest dynamics under global change.** Improved representation of vertical variation in forest canopy architecture and ecosystem function is critical for representing thermal sensitivity and has repeatedly been identified as an important step toward reduced uncertainty and more accurate characterization of biologically mediated feedbacks [@moorcroftMethodScalingVegetation2001; @banerjeeEffectVerticalCanopy2018; @bonanMovingIncorrectUseful2021]. In addition, this review finds that, although larger trees face disproportionate stress during drought, understory trees have fewer cooling mechanisms and may also be stifled under hot, humid conditions. **A key question is whether existing models adequately represent the processes that underpin understory tree responses to thermal stress as well as large tree responses.** Most models have been developed to capture dynamics in the canopy, or overstory, given the disproportionate role of these large trees in ecosystem dynamics. Less attention has focused on developing and validating understory tree dynamics and responses to perturbations in models. This is likely due in large part to the fact that observations are only beginning to shed light on the vulnerability of understory trees to changes in temperature and other abiotic conditions [e.g., @rollinsonClimateSensitivityUnderstory2020].

Though an improvement over big-leaf models, DGVMs that separate the canopy into only two layers (e.g., sunlit and shaded portions) may not be able to capture important within-canopy variation in terms of leaf dynamics [e.g., seasonal shifts in vertical leaf area distributions, @smithSeasonalDroughtrelatedChanges2019a] and functions (e.g., thermal responses, as we present in this paper). At the very least, multi-layered ecosystem models will likely be necessary for accurately predicting future forest function [e.g., @depurySimpleScalingPhotosynthesis1997; @bonanMovingIncorrectUseful2021]. In addition, capturing the vertical gradients in NPP, GPP, and respiration, and subsequently the net ecosystem effects, requires improved characterization of the functional response of leaf-level processes to vertically varying abiotic conditions, and the role of traits and abiotic conditions in mediating responses to thermal sensitivity.

@@

Detailed structural information from lidar data has been used to measure tree height, DBH, and crown dimensions [@fisherECOSTRESSNASANext2020], evaluate spatial variation in vertical leaf area density profiles [@dettoSpatialVariabilityTropical2015], and resolve complex seasonal and diurnal variation in shortwave radiation forcing in ecosystems [@musselmanEstimationSolarDirect2013]. In addition, shading by the overstory can be better represented to predict understory temperatures by modeling the time-varying interaction of the forest canopy with solar radiation, as has been done in other LIDAR-based analyses [e.g., @davisLiDARderivedTopographyForest2019]. Satellite and airborne thermal infrared remote sensing analyses are now being conducted at regional and continental scales, although very few applications exist at finer-scales necessary for understanding of vertical variation in plant canopy temperatures [@johnstonWhatLiesVertical2020 in review]. [@pauTropicalForestTemperature2018] used data from a tower-based FLIR camera in combination with eddy-covariance data and found that tropical forest canopy temperatures were more strongly associated with GPP than air temperatures or VPD. In a savanna system in the western U.S., [@johnstonWhatLiesVertical2020 in review] found significant vertical variation in plant temperature, with warmer temperatures in the understory compared to the top of the blue oak canopies. This finding was the opposite of what was expected and occurred in association with very high grass temperatures. The growing availability of these data makes this an opportune time to link fine-scale and landscape-scale measurements with the type of in situ measurements reviewed above to further explore ecosystem-scale patterns in vertical temperature gradients seasonally and across biomes.
